\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Breast Cancer Prediction Using Machine Learning Algorithms}

\author{
\IEEEauthorblockN{Sarah Sejoro}
\IEEEauthorblockA{\textit{Jiann-Ping Hsu College of Public Health}\\
\textit{Georgia Southern University}\\
Statesboro, GA, United States\\
ss81506@georgiasouthern.edu}
\and
\IEEEauthorblockN{Anandhi Kandaswamy}
\IEEEauthorblockA{\textit{Jack N. Averitt College of Graduate Studies}\\
\textit{Georgia Southern University}\\
Statesboro, GA, United States\\
ak20427@georgiasouthern.edu}
\and
\IEEEauthorblockN{Bright Agbenu}
\IEEEauthorblockA{\textit{Jiann-Ping Hsu College of Public Health}\\
\textit{Georgia Southern University}\\
Statesboro, GA, United States\\
ba09534@georgiasouthern.edu}
}

\maketitle

\begin{IEEEkeywords}
Breast cancer, machine learning, disease classification, diagnostic prediction, feature engineering
\end{IEEEkeywords}

\section{Introduction}
Cancer remains a major public health burden, with breast cancer representing a large share of diagnoses among women in the United States. Female breast cancer accounts for about 15.5\% of all new cancer cases in the U.S., and the National Cancer Institute estimates 316,950 new female breast cancer cases in 2025, with a 5-year relative survival of 91.7\% \cite{b1,b2}. Women face an estimated lifetime risk of about 1 in 8 ($\approx$13\%) \cite{b3}, while male breast cancer is rare but still occurs \cite{b4}. Because outcomes depend strongly on early detection and appropriate treatment, improving diagnostic accuracy and consistency is a continuing priority in both clinical practice and research. At the same time, traditional screening and diagnostic pathways can involve false positives, follow-up biopsies, and variable performance across settings, which motivates complementary decision-support approaches \cite{b5,b6}.

Recent advances in digital pathology, machine learning (ML), and deep learning (DL) have expanded the toolkit for breast cancer prediction, ranging from benign versus malignant classification using structured nuclear features to imaging-driven and radiomics-based prediction tasks \cite{b7,b8,b9}. In our team project, the focus is supervised learning for breast cancer classification using a reproducible pipeline and an ensemble of strong baseline models, motivated by prior findings that performance depends as much on data preparation and validation design as it does on the choice of algorithm \cite{b10,b11}.

\section{Methodology}
This literature review followed a structured selection process centered on supervised learning methods for breast cancer prediction and closely related disease-classification workflows. Search terms included combinations of ``breast cancer prediction,'' ``breast cancer diagnosis,'' ``supervised learning,'' ``machine learning classification,'' ``Wisconsin Diagnostic Breast Cancer,'' ``feature selection,'' ``random forest,'' ``support vector machine,'' ``logistic regression,'' ``gradient boosting,'' ``radiomics,'' and ``digital pathology.'' Database searching emphasized PubMed/PubMed Central, ScienceDirect, and Google Scholar, with targeted retrieval from journal and repository pages when needed for dataset documentation and publication metadata \cite{b2,b12}. Articles were included if they (1) applied ML/DL to breast cancer prediction or closely related diagnostic classification tasks, (2) described the data type and modeling workflow clearly enough to interpret the approach, and (3) reported evaluation outcomes using standard metrics (e.g., accuracy, AUC, sensitivity/specificity, or related measures). Articles were excluded if they lacked methodological detail, did not address prediction/classification, or were primarily conceptual without empirical evaluation. Because our project uses structured diagnostic features rather than raw imaging, imaging-heavy studies were used mainly to contextualize broader trends and validation challenges rather than to define the project's core modeling pipeline \cite{b7,b8}.

\section{Literature Review}
Breast cancer prediction is not a single task in the literature. It includes benign versus malignant classification from structured diagnostic features, prediction of treatment response using radiomics, prognosis modeling using time-to-event outcomes, and prediction of lymph node involvement using imaging-derived and clinicopathologic variables \cite{b14,b15,b16}. Despite these varied endpoints, a consistent theme is that supervised learning is effective when features encode clinically meaningful structure and when evaluation procedures reflect the intended use case \cite{b7,b10}.

For structured, tabular breast cancer datasets, traditional supervised learners remain central. Reviews and comparative studies repeatedly highlight logistic regression, Naive Bayes, support vector machines, decision trees, random forests, and boosting methods as strong baselines, especially when paired with disciplined preprocessing and feature selection \cite{b7,b10,b9}. This framing aligns directly with the Wisconsin Diagnostic Breast Cancer (WDBC) dataset, which contains features computed from digitized fine-needle aspirate images that describe cell nuclei characteristics and supports benign/malignant classification \cite{b12}. Studies using WDBC-style settings commonly report competitive performance across multiple supervised models, reinforcing the idea that careful benchmarking is often more informative than relying on a single ``best'' algorithm \cite{b11,b18}.

Within breast-cancer-specific ML research, feature engineering and feature selection repeatedly emerge as performance drivers. Work discussing nuclear feature sets emphasizes that real-valued measurements such as radius, perimeter, and concavity can support classification, and broader reviews describe feature prioritization as a practical way to reduce model complexity while preserving diagnostic signal \cite{b17,b9}. Consistent with this, La Moglia and Almustafa report that logistic regression and gradient boosting variants can achieve strong test performance in breast cancer classification settings, with performance patterns shifting depending on whether feature selection is applied \cite{b13}. Beyond tumor classification, Bruny\'e et al.\ extend the ML diagnostic frame by showing that models can predict diagnostic accuracy based on pathologists' viewing behavior during breast biopsy interpretation, reinforcing that evaluation and workflow context matter alongside raw predictive performance \cite{b19}.

At the same time, several limitations recur across studies. First, results are often difficult to compare due to heterogeneous datasets, inconsistent preprocessing and feature extraction choices, and differences in reported metrics, which can inflate confidence in ``best model'' claims when pipelines are not fully documented \cite{b10,b8}. Second, external validity remains a persistent concern: models may perform well on internal splits but degrade when applied across institutions, scanners, or patient populations, which is particularly visible in radiomics and prognosis contexts \cite{b14,b15,b16}. Third, interpretability and clinical integration are frequently underdeveloped relative to performance reporting; high accuracy or AUC alone does not guarantee that a model's errors, uncertainty, or subgroup behavior are clinically acceptable \cite{b7,b8}.

\section{Synthesis and Conclusion}
Across the reviewed literature, the strongest consensus is that breast cancer prediction benefits most from transparent pipelines, clinically sensible feature representations, and evaluation designs that match the endpoint being predicted. Traditional supervised learning models and ensembles remain strong choices for structured feature sets like WDBC, while radiomics and prognosis studies illustrate that supervised learners can extend to more complex prediction targets when feature extraction and validation are carefully handled \cite{b11,b18,b14,b15}. Comparative benchmarking work also reinforces that no single classifier is universally best across medical datasets, which supports a model-comparison approach rather than a one-model narrative \cite{b10}.

The most relevant gaps for our project are practical and reproducible: inconsistent reporting of preprocessing and feature selection decisions, limited attention to class-sensitive metrics beyond accuracy, and limited interpretability and error analysis to explain what drives model performance and where models fail \cite{b7,b8}. Our project addresses these gaps by implementing a reproducible supervised-learning workflow on the WDBC dataset, comparing multiple baseline classifiers and an ensemble, using a clearly separated validation and test strategy, and reporting class-sensitive metrics alongside confusion-pattern error analysis and interpretable outputs such as feature importance for clinically meaningful discussion \cite{b12,b11,b18}.

\section*{}
\begin{thebibliography}{00}

\bibitem{b1} J. Makki, ``Diversity of breast carcinoma: Histological subtypes and clinical relevance,'' \emph{Clinical Medicine Insights: Pathology}, vol. 8, pp. 23--31, 2015, doi: 10.4137/CPath.S31563.

\bibitem{b2} National Cancer Institute, ``Cancer Stat Facts: Female Breast Cancer,'' SEER, 2025. [Online]. Available: \url{https://seer.cancer.gov/statfacts/html/breast.html}. Accessed: Feb. 20, 2026.

\bibitem{b3} American Cancer Society, ``Breast Cancer Statistics: How Common Is Breast Cancer?,'' 2026. [Online]. Available: \url{https://www.cancer.org/cancer/types/breast-cancer/about/how-common-is-breast-cancer.html}. Accessed: Feb. 20, 2026.

\bibitem{b4} American Cancer Society, ``Key Statistics for Breast Cancer in Men,'' 2026. [Online]. Available: \url{https://www.cancer.org/cancer/types/breast-cancer-in-men/key-statistics.html}. Accessed: Feb. 20, 2026.

\bibitem{b5} P. Jaglan, R. Dass, and M. Duhan, ``Breast cancer detection techniques: Issues and challenges,'' \emph{J. Institution of Engineers (India): Series B}, vol. 100, no. 4, pp. 379--386, 2019, doi: 10.1007/s40031-019-00391-2.

\bibitem{b6} C. Shang and D. Xu, ``Epidemiology of breast cancer,'' \emph{Oncologie}, vol. 24, no. 4, pp. 649--663, 2022, doi: 10.32604/oncologie.2022.027640.

\bibitem{b7} M. M. Ahsan, S. A. Luna, and Z. Siddique, ``Machine-learning-based disease diagnosis: A comprehensive review,'' \emph{Healthcare}, vol. 10, no. 3, art. 541, 2022, doi: 10.3390/healthcare10030541.

\bibitem{b8} A. Jafari, ``Machine-learning methods in detecting breast cancer and related therapeutic issues: A review,'' \emph{Computer Methods in Biomechanics and Biomedical Engineering: Imaging \& Visualization}, vol. 12, no. 1, art. 2299093, 2024, doi: 10.1080/21681163.2023.2299093.

\bibitem{b9} W. Yue, Z. Wang, H. Chen, A. Payne, and X. Liu, ``Machine learning with applications in breast cancer diagnosis and prognosis,'' \emph{Designs}, vol. 2, no. 2, art. 13, 2018, doi: 10.3390/designs2020013.

\bibitem{b10} M. A. Moreno-Ibarra, Y. Villuendas-Rey, D. L\'opez-S\'anchez, and J. H. Arroyo-N\'u\~nez, ``Classification of diseases using machine learning algorithms: A comparative study,'' \emph{Mathematics}, vol. 9, no. 15, art. 1817, 2021, doi: 10.3390/math9151817.

\bibitem{b11} N. Al-Azzam and I. Shatnawi, ``Comparing supervised and semi-supervised machine learning models on diagnosing breast cancer,'' \emph{Annals of Medicine and Surgery}, vol. 62, pp. 53--64, 2021, doi: 10.1016/j.amsu.2020.12.043.

\bibitem{b12} W. Wolberg, O. Mangasarian, N. Street, and W. Street, ``Breast Cancer Wisconsin (Diagnostic)'' [Dataset], UCI Machine Learning Repository, 1993, doi: 10.24432/C5DW2B.

\bibitem{b13} A. La Moglia and K. M. Mohamad Almustafa, ``Breast cancer prediction using machine learning classification algorithms,'' \emph{Intelligence-Based Medicine}, vol. 11, art. 100193, 2025, doi: 10.1016/j.ibmed.2024.100193.

\bibitem{b14} X. Li, C. Li, H. Wang, L. Jiang, and M. Chen, ``Comparison of radiomics-based machine-learning classifiers for the pretreatment prediction of pathologic complete response to neoadjuvant therapy in breast cancer,'' \emph{PeerJ}, vol. 12, e17683, 2024, doi: 10.7717/peerj.17683.

\bibitem{b15} J. Xiao, M. Mo, Z. Wang, C. Zhou, J. Shen, J. Yuan, Y. He, and Y. Zheng, ``The application and comparison of machine learning models for the prediction of breast cancer prognosis: Retrospective cohort study,'' \emph{JMIR Medical Informatics}, vol. 10, no. 2, e33440, 2022, doi: 10.2196/33440.

\bibitem{b16} J. Yao, W. Zhou, S. Xu, X. Jia, J. Zhou, X. Chen, and W. Zhan, ``Machine learning-based breast tumor ultrasound radiomics for pre-operative prediction of axillary sentinel lymph node metastasis burden in early-stage invasive breast cancer,'' \emph{Ultrasound in Medicine \& Biology}, vol. 50, no. 2, pp. 229--236, 2024, doi: 10.1016/j.ultrasmedbio.2023.10.004.

\bibitem{b17} C. G. Yedjou, S. S. Tchounwou, R. A. Al\'o, R. Elhag, B. K. Mochona, and L. Latinwo, ``Application of machine learning algorithms in breast cancer diagnosis and classification,'' \emph{BioMed Research International}, 2021, art. 9983652.

\bibitem{b18} S. Zhou, C. Hu, S. Wei, and X. Yan, ``Breast cancer prediction based on multiple machine learning algorithms,'' \emph{Technology in Cancer Research \& Treatment}, vol. 23, 2024, doi: 10.1177/15330338241234791.

\bibitem{b19} T. T. Bruny\'e, K. Booth, D. Hendel, K. F. Kerr, H. Shucard, D. L. Weaver, and J. G. Elmore, ``Machine learning classification of diagnostic accuracy in pathologists interpreting breast biopsies,'' \emph{Journal of the American Medical Informatics Association}, vol. 31, no. 3, pp. 552--562, 2024, doi: 10.1093/jamia/ocad232.

\end{thebibliography}

\end{document}

